---
title: "Bayesian Reanalysis of METR AI Developer Productivity Study"
format:
  html:
    embed-resources: true
---

```{r}
#| label: setup
#| message: false

util <- new.env()
source("mcmc_visualization_tools/r/mcmc_analysis_tools_rstan.R", local = util)
source("mcmc_visualization_tools/r/mcmc_visualization_tools.R", local = util)

library(rstan)
rstan_options(auto_write = TRUE)
options(mc.cores = parallel::detectCores())

```

# Exploratory Data Analysis

```{r}
#| label: load-data

dat <- read.csv("data/data_complete.csv")

# Compute total implementation time (minutes)
dat$total_time <- dat$initial_implementation_time + dat$post_review_implementation_time

# Forecast time (minutes)
dat$forecast <- dat$predicted_time_no_ai

# Recode treatment: 1 = AI access, 0 = AI restricted
dat$ai_access <- 1 - dat$ai_treatment
```

```{r}
#| label: fig-completion-times
#| fig-cap: "Distribution of observed completion times (minutes)"

util$plot_line_hist(dat$total_time[!is.na(dat$total_time)],
                    0, 1500, 50,
                    xlab = "Completion time (minutes)")
```

```{r}
#| label: fig-forecast-times
#| fig-cap: "Distribution of forecast times (minutes)"

util$plot_line_hist(dat$forecast,
                    0, 800, 25,
                    xlab = "Forecast time (minutes)")
```

```{r}
#| label: data-summary

cat("N (complete cases):", sum(!is.na(dat$total_time)), "\n")
cat("N (missing):", sum(is.na(dat$total_time)), "\n")
cat("\n")

cat("total_time (minutes):\n")
cat("  median:", median(dat$total_time, na.rm = TRUE), "\n")
cat("  mean:", round(mean(dat$total_time, na.rm = TRUE), 1), "\n")
cat("  sd:", round(sd(dat$total_time, na.rm = TRUE), 1), "\n")
cat("  range:", range(dat$total_time, na.rm = TRUE), "\n")

cat("\nforecast (minutes):\n")
cat("  median:", median(dat$forecast), "\n")
cat("  mean:", round(mean(dat$forecast), 1), "\n")
cat("  sd:", round(sd(dat$forecast), 1), "\n")
cat("  range:", range(dat$forecast), "\n")

cat("\nai_access:\n")
cat("  n(AI allowed):", sum(dat$ai_access == 1), "\n")
cat("  n(AI restricted):", sum(dat$ai_access == 0), "\n")
```

# Model 1: Bayesian Regression

## Model Structure

We model completion time with a lognormal likelihood, regressing on treatment and forecast:

$$y_i \sim \text{Lognormal}(\mu_i, \sigma)$$ $$\mu_i = \alpha + \beta_{\text{trt}} \cdot \text{ai_access}_i + \beta_{\text{forecast}} \cdot \log(\text{forecast}_i / 90)$$

The baseline forecast is set to 90 minutes (approximately the median). This means:

-   $\alpha$ = expected log(completion time) when forecast = 90 min and AI restricted
-   $\exp(\alpha)$ = median completion time at baseline
-   $\beta_{\text{trt}}$ = effect of AI access on log completion time
-   $\beta_{\text{forecast}}$ = elasticity (if forecast doubles, completion time multiplies by $2^{\beta_{\text{forecast}}}$)

## Prior Development

We calibrate priors using domain expertise about plausible parameter ranges. For each parameter, we identify extremity thresholds and set the prior SD so that 1% of probability falls beyond each threshold.

**Intercept (**$\alpha$): At baseline (forecast = 90 min, AI restricted), median completion time plausibly ranges from 45 to 180 minutes. On log scale: $\log(45) \approx 3.8$ to $\log(180) \approx 5.2$, centered at $\log(90) \approx 4.5$.

$$\alpha \sim \text{Normal}(\log(90), 0.30)$$

**Treatment effect (**$\beta_{\text{trt}}$): A 95% reduction (log change of -3) or 5x increase (log change of +1.6) from AI access would be implausible.

$$\beta_{\text{trt}} \sim \text{Normal}(0, 0.7)$$

**Forecast elasticity (**$\beta_{\text{forecast}}$): We expect forecasts to be predictive. An elasticity between 0.25 and 1.75 is plausible; values outside this range are unlikely.

$$\beta_{\text{forecast}} \sim \text{Normal}(1, 0.32)$$

**Residual dispersion (**$\sigma$): Starting with $\sigma = 1$ and tuning based on prior predictive checks. Goal: 99% of simulated completion times between 10 min and 24 hours.

## Prior Predictive Check

We simulate data from the prior to verify our priors produce plausible datasets before seeing any real data.

```{r}
#| label: model1-prior-data

complete_idx <- !is.na(dat$total_time)

stan_data <- list(
  N = sum(complete_idx),
  forecast = dat$forecast[complete_idx],
  ai_access = dat$ai_access[complete_idx],
  x0 = 90
)
```

```{r}
#| label: model1-prior-fit

fit_prior <- stan(
  file = "stan_programs/model1_prior.stan",
  data = stan_data,
  algorithm = "Fixed_param",
  iter = 1000,
  chains = 1,
  seed = 1234
)

samples_prior <- util$extract_expectand_vals(fit_prior)
```

### Prior on $\alpha$ (log baseline median)

Dashed lines show the extremity thresholds: log(45) and log(180) minutes.

```{r}
#| label: fig-prior-alpha
#| fig-cap: "Prior on alpha: log(median completion time) at baseline"

util$plot_line_hist(samples_prior$alpha, 3.5, 5.5, 0.1,
                    xlab = "alpha (log minutes)")
abline(v = log(45), lty = 2, lwd = 2)   # Lower threshold: 45 min
abline(v = log(180), lty = 2, lwd = 2)  # Upper threshold: 180 min
```

### Prior on $\mu$ (log median completion time)

Distribution of $\mu = \alpha + \beta_{\text{trt}} \cdot \text{ai_access} + \beta_{\text{forecast}} \cdot \log(\text{forecast}/90)$ across all observations and prior draws. Dashed lines show log(10 min) and log(24 hours) — the target range for plausible completion times.

```{r}
#| label: fig-prior-mu
#| fig-cap: "Prior predictive: log median completion time (with treatment effect)"

util$plot_hist_quantiles(samples_prior, 'mu', 0, 8, 0.25,
                         xlab = "mu (log minutes)")
abline(v = log(10), lty = 2, lwd = 2)    # 10 min
abline(v = log(1440), lty = 2, lwd = 2)  # 24 hours
```

```{r}
#| label: fig-prior-mu-no-trt
#| fig-cap: "Prior predictive: log median completion time (no treatment effect)"

util$plot_hist_quantiles(samples_prior, 'mu_no_trt', 0, 8, 0.25,
                         xlab = "mu (log minutes)")
abline(v = log(10), lty = 2, lwd = 2)    # 10 min
abline(v = log(1440), lty = 2, lwd = 2)  # 24 hours
```

### Prior predictive on $y$ (completion time in minutes)

The lognormal model adds dispersion $\sigma$ around the median. Dashed lines show the target range: 10 min to 24 hours (1440 min).

```{r}
#| label: fig-prior-y
#| fig-cap: "Prior predictive: completion time (minutes)"

util$plot_hist_quantiles(samples_prior, 'y_sim', 0, 1500, 25,
                         xlab = "Completion time (minutes)")
abline(v = 10, lty = 2, lwd = 2)    # 10 min
abline(v = 1440, lty = 2, lwd = 2)  # 24 hours
```

```{r}
#| label: fig-prior-y-no-trt
#| fig-cap: "Prior predictive: completion time without treatment (minutes)"

util$plot_hist_quantiles(samples_prior, 'y_sim_no_trt', 0, 1500, 25,
                         xlab = "Completion time (minutes)")
abline(v = 10, lty = 2, lwd = 2)    # 10 min
abline(v = 1440, lty = 2, lwd = 2)  # 24 hours
```

### Prior predictive at baseline (forecast = 90 min, no treatment)

Dashed lines show the alpha extremity thresholds on the original scale: 45 and 180 minutes.

```{r}
#| label: fig-prior-y-baseline
#| fig-cap: "Prior predictive: completion time at baseline"

util$plot_line_hist(samples_prior$y_sim_baseline, 0, 600, 20,
                    xlab = "Completion time (minutes)")
abline(v = 45, lty = 2, lwd = 2)   # Lower threshold
abline(v = 180, lty = 2, lwd = 2)  # Upper threshold
```

### Prior predictive on log scale

Dashed lines show log(10 min) and log(24 hours).

```{r}
#| label: fig-prior-log-y
#| fig-cap: "Prior predictive: log completion time"

util$plot_hist_quantiles(samples_prior, 'log_y_sim', 0, 10, 0.25,
                         xlab = "log(completion time)")
abline(v = log(10), lty = 2, lwd = 2)    # 10 min
abline(v = log(1440), lty = 2, lwd = 2)  # 24 hours
```

### Prior on treatment effect (percentage lift)

Since $\beta_{\text{trt}}$ is the treatment effect on the log scale, the percentage change in completion time from AI access is $(e^{\beta_{\text{trt}}} - 1) \times 100\%$. Negative values indicate speedup; positive values indicate slowdown. Dashed lines show the extremity thresholds: 95% reduction (-95%) and 5x increase (+400%).

```{r}
#| label: fig-prior-ate-pct
#| fig-cap: "Prior on treatment effect: percentage change in completion time"

pct_lift <- (exp(samples_prior$beta_trt) - 1) * 100
util$plot_line_hist(pct_lift, -100, 500, 20,
                    xlab = "Treatment effect (% change)")
abline(v = 0, lty = 1, lwd = 1)          # No effect
abline(v = -95, lty = 2, lwd = 2)        # 95% reduction (implausible)
abline(v = 400, lty = 2, lwd = 2)        # 5x increase (implausible)
```

## Model Fitting

```{r}
#| label: model1-fit-data

complete_idx <- !is.na(dat$total_time)

stan_data_fit <- list(
  N = sum(complete_idx),
  y = dat$total_time[complete_idx],
  forecast = dat$forecast[complete_idx],
  ai_access = dat$ai_access[complete_idx],
  x0 = 90
)
```

```{r}
#| label: model1-fit

fit <- stan(
  file = "stan_programs/model1.stan",
  data = stan_data_fit,
  seed = 8877273,
  refresh = 100
)
```

```{r}
#| label: model1-diagnostics

util$check_all_hmc_diagnostics(util$extract_hmc_diagnostics(fit))
samples <- util$extract_expectand_vals(fit)
util$check_all_expectand_diagnostics(samples)
```

## Posterior Retrodictive Checks

### Log completion time

Comparing posterior predictive distribution of log completion times against observed data (black histogram).

```{r}
#| label: fig-retro-log-y
#| fig-cap: "Posterior retrodictive: log completion time"

util$plot_hist_quantiles(samples, 'log_y_rep', 2, 8, 0.25,
                         baseline_values = log(stan_data_fit$y),
                         xlab = "log(completion time)")
```

### Treatment effect (percentage lift)

Posterior distribution of the treatment effect as percentage change in completion time. The dashed line shows the observed difference in median completion times between AI-allowed and AI-restricted groups.

```{r}
#| label: fig-retro-ate-pct
#| fig-cap: "Posterior: treatment effect (% change in completion time)"

# Observed percentage difference in medians
y_ai <- dat$total_time[complete_idx & dat$ai_access == 1]
y_no_ai <- dat$total_time[complete_idx & dat$ai_access == 0]
observed_pct_diff <- (median(y_ai) / median(y_no_ai) - 1) * 100

util$plot_line_hist(samples$pct_lift, -80, 80, 5,
                    xlab = "Treatment effect (% change)")
abline(v = 0, lty = 1, lwd = 1)
abline(v = observed_pct_diff, lty = 2, lwd = 2)
```

### Parameter summaries

```{r}
#| label: model1-summary

print(fit, pars = c("alpha", "beta_trt", "beta_forecast", "sigma", "pct_lift"))
```

### Conditional mean by task exposure

```{r}
#| label: fig-retro-cond-exposure
#| fig-cap: "Posterior retrodictive: completion time by prior task exposure"

# Get complete case data (what was modeled)
dat_complete <- dat[complete_idx, ]

# Find which complete cases have non-NA exposure
has_exposure <- !is.na(dat_complete$prior_task_exposure_1_to_5)

# Indices into y_rep (1:N for complete cases)
n_cond_mean <- which(has_exposure)

# Covariate values and observed outcomes for these observations
obs_xs <- dat_complete$prior_task_exposure_1_to_5[has_exposure]
obs_ys <- dat_complete$total_time[has_exposure]

# y_rep names for these observations
names <- paste0('y_rep[', n_cond_mean, ']')

util$plot_conditional_mean_quantiles(samples, names, obs_xs,
                                     0.5, 5.5, 1, obs_ys,
                                     xlab = "Prior task exposure (1-5)",
                                     ylab = "Completion time (minutes)")
```

Conditional Means by Resource Needs

```{r}
#| label: fig-retro-cond-exposure
#| fig-cap: "Posterior retrodictive: completion time by resource needs"

# Get complete case data (what was modeled)
dat_complete <- dat[complete_idx, ]

# Find which complete cases have non-NA exposure
has_resource_needs <- !is.na(dat_complete$external_resource_needs_1_to_3)

# Indices into y_rep (1:N for complete cases)
n_cond_mean <- which(has_resource_needs)

# Covariate values and observed outcomes for these observations
obs_xs <- dat_complete$external_resource_needs_1_to_3[has_resource_needs]
obs_xs[obs_xs == 5] <- 3
obs_ys <- dat_complete$total_time[has_resource_needs]

# y_rep names for these observations
names <- paste0('y_rep[', n_cond_mean, ']')

util$plot_conditional_mean_quantiles(samples, names, obs_xs,
                                     0.5, 3.5, 1, obs_ys,
                                     xlab = "Resource Needs (1 to 3)",
                                     ylab = "Completion time (minutes)")
```
